A free/open-source machine translation system for English to Kazakh

### 2015-03-02 engkaz-eamt-2015/engkaz-eamt-2015.tex
fixes
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@59143 72bbbca6-d526-0410-a7d9-f06f51895060
fixes
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@59182 72bbbca6-d526-0410-a7d9-f06f51895060

This paper presents a shallow-transfer rule-based machine translation system from English into Kazakh language. First of all, English language is a West Germanic language and Kazakh belong to group of Turkic languages. Therefore, the complex agglutinative morphology of Kazakh language is very different from morphologically not too complex language like English; on the other hand, Turkic language morphology shows clear morphotactics (ordering of morphemes), its morphophonology shows complex phonological changes to due to interactions between neighboring morphemes (vowel harmony, sonorization, etc.) many of which are explicitly represented in writing.
On the other hand, there are many differences between the syntax of Turkic languages and English:.subject-adverb–object–verb order (compare subject–verb–object-adverb in English), use of postpositions (compare prepositions in English), head-final syntax with modifiers and specifiers always preceding the modified/specified (normally following in English), overt case marking allowing for a rather free ordering of arguments (versus a more fixed order in English), lack of definite articles (extensively used in English), verbal-noun-centered structures where English uses modal verbs (must, have to, want to) or verbal-noun or verbal-adjective-centered constructions where English has subordinate clauses using finite verbs with relatives or subordinating conjunctions (the book which I read, the place where I saw him, before he came), lack of a parallel of the English verb have, as used for possession, etc. For an account (in Russian) of syntax differences between English and Kazakh, see Печерских \& Амангельдина (2012).
This paper describes work in progress of development of machine translation system for English-to-Kazakh, which developed by using Apertium free/open-source machine translation platform (Forcada et al. 2011, http://www.apertium.org).

@@

\begin{table}
  \begin{tabular}{|l|r|r|}
    \hline
    \textbf{System} & \textbf{BLEU} & \textbf{WER} \\
    \hline
    \texttt{apertium-eng-kaz} & 44.23 & 42.88 \\
    Google & 62.00 & 25.92 \\
    Sanasoft & 3.09 & 99.72 \\
    Trident & 5.06 & 92.86 \\
    \hline
  \end{tabular}
\end{table}

@@

\com{task-based evaluation (WER, assim, TM-patching?) -FMT}

\com{Compare with GT + other systems -FMT}

@@

The current English-Kazakh machine translation system already successfully translates noun-phrases, verb-phrases, prepositional-phrases, and adjectival-phrases, and contains a good vocabulary for testing purposes.

@@

>>>>

This paper presents a shallow-transfer rule-based machine translation system from English into Kazakh language. First of all, English language is a West Germanic language and Kazakh belong to group of Turkic languages. Therefore, the complex agglutinative morphology of Kazakh language is very different from morphologically not too complex language like English; on the other hand, Turkic language morphology shows clear morphotactics (ordering of morphemes), its morphophonology shows complex phonological changes to due to interactions between neighboring morphemes (vowel harmony, sonorization, etc.) many of which are explicitly represented in writing.

On the other hand, there are many differences between the syntax of Turkic languages and English:.subject-adverb–object–verb order (compare subject–verb–object-adverb in English), use of postpositions (compare prepositions in English), head-final syntax with modifiers and specifiers always preceding the modified/specified (normally following in English), overt case marking allowing for a rather free ordering of arguments (versus a more fixed order in English), lack of definite articles (extensively used in English), verbal-noun-centered structures where English uses modal verbs (must, have to, want to) or verbal-noun or verbal-adjective-centered constructions where English has subordinate clauses using finite verbs with relatives or subordinating conjunctions (the book which I read, the place where I saw him, before he came), lack of a parallel of the English verb have, as used for possession, etc. For an account (in Russian) of syntax differences between English and Kazakh, see Печерских \& Амангельдина (2012).

This paper describes work in progress of development of machine translation system for English-to-Kazakh, which developed by using Apertium free/open-source machine translation platform (Forcada et al. 2011, http://www.apertium.org).

@@

\begin{table}
  \begin{tabular}{|l|r|r|}
    \hline
    \textbf{System} & \textbf{BLEU} & \textbf{WER} \\
    \hline
    \texttt{apertium-eng-kaz} & 44.23 & 42.88 \\
    Google & 62.00 & 25.92 \\
    Sanasoft & 21.00 & 74.52 \\
    \hline
  \end{tabular}
\end{table}

@@

As can be seen,  coverage of dictionary is good, average score is 94\%. Our system outperforms other available RBMT system, but fall short of the state-of-the-art performance. However, unlike the state of the art, which uses unavailable corpora, our system is free and open and can be improved.

MT systems have some mistakes in translations, which was shown in <...> brackets. Google MT system has BLEU score higher, but in translation separate phrases, it has common error: do not assign right possessive and case. Sanasoft system has more errors with choosing not right translations or do not translate word from English.

@@

We have presented the rule-based MT system from English to Kazakh. The current English-Kazakh machine translation system already successfully translates noun-phrases, verb-phrases, prepositional-phrases, and adjectival-phrases, and contains a good vocabulary for testing purposes. 
We plan to continue developing English-Kazakh pair, going to extend the coverage to 98% on the corpora, which we used before. Additionally, we will improve the quality of translation by adding more rules, such like constraint grammar rules, structural transfer and lexical rules. The future plan is to use the created data with other open-source MT system with Turkic languages or system with Kazakh language on right side to make transfer systems between the Turkic or other language pairs. Related work is currently ongoing with Russian-Kazakh. 
Our system is available as free/open-source software and the whole system may be downloaded from sourceforge(apertium-eng-kaz1, apertium-kaz2).

@@

MLF thanks the Kazakh state program for the attraction of foreign scholars and Prof. Ualsher Tukeyev for supporting his visit to the Kazakh National University, where part of this work was carried out. We also thank Prof. Tukeyev for his valuable input.
